---
title: "Violence and Immigration to the United States"
author: "Matt Aadland"
date: "Tuesday, April 19, 2016"
output: html_document
---

### Impact Statement
One strategic implication of the efficacy of security forces in El Salvador, Guatemala, and Honduras is the impact on immigration to the United States. It is possible that insufficient physical security provided by local and national security forces within the three nations directly influences the decision for migrants to flee to the United States.  In this piece, we analyze this statistical relationship between physical security in the three nations that comprise the Northern Triangle and migration to the United States. 

### Bottom Line Upfront:
Based on the analysis below, we argue that a statistical relationship exists between increased violence in the Northern Triangle and increased migration to the United States.  This means that the security situation in El Salvador, Guatemala, and Honduras is highly relevant to the United States interest and immigration policy. The results of our model can be used to anticipate increased migration to the United States based on reports of homicides inside of the Northern Triangle.   

The rest of this piece is organized as follows: first, we detail the sources and methods we used in our analysis, then we display the software commands we used to generate our results, finally we end the piece with our interpretation and a visualization showing the relationship between security in the Northern Triangle and migration to the United States. 


### Sources and Methods: 
We use a linear regression analysis to test the impact that security conditions in the Northern Triangle have on migration rates to the United States. Regression provides a simple and powerful tool to examine whether or not increased violence will lead to increased migration to the United States. 

To test the relationship, we run a regression analysis of migration as the response variable and the physical security of the Northern Triangle as the predictor variable. We also control for confounding variables related to economic security and population growth. We use a statistical test of significance on the slope of the predictor variable as a measure of its relevance to determining migration rates. 

### Response Variable:
To test the impact that security conditions have in the Northern Triangle on migration to the United States, we use the enforcement data from the US Office of Immigration Statistics. The data we use as a proxy to the total migration to the United States are the enforcement actions taken by Department of Homeland Security. Specifically, we measure the number of aliens apprehended by nationality. The DHS records enforcement actions to include foreign nationals who are "inadmissible, apprehended, removed or returned for violating the Immigration and Nationality Act" (Office of Immigration Statistics 2014).  Data are available for fiscal years 2002 through 2013. 

The number of alien apprehensions is an imperfect measure for total migration because it introduces extraneous variables for different enforcement policies by year as well as failing to measure unsuccessful apprehensions. It also fails to measure an increase in legal migration from the target country. While these factors make it difficult to develop a predictive model of migration, it is still a useful measure for trends in migration and endogenous factors influencing outmigration. 

### Predictor Variable:
To measure the security conditions in our target countries, we use the United Nations Office on Drugs and Crime Intentional Homicide Statistics dataset. Intentional homicides estimates of unlawful homicides" are purposely inflicted as a result of domestic disputes, interpersonal violence, and violent conflicts over land resources, intergang violence over turf or control, and predatory violence and killing by armed groups" (World Development Indicators, The World Bank 2016).  This definition excludes killings committed by cohesive groups in the case of armed conflicts. The data are collected off of the World Development Indicators API (World Bank 2016). 


### Control Variables: 
It is useful to measure the economic conditions of El Salvador, Guatemala, and Honduras as it relates to migration policy. It is possible that a deteriorating economic environment increases migration to the United States rather than a fragile security situation. To control for confounding economic variables, we include variables for gross national product growth per year as well as gross national income. It is also possible that population growth helps drive increased migration rates. To control for this, we introduce a third control variable for population growth. We also normalize our response variable for  population by dividing the total number of alien apprehensions by the population for that country year. All of the control variables were collected from the World Development Indicators API (World Bank 2016).

### Data Preparation:
```{r,echo=FALSE}
setwd("C:/Users/aadlandma/Desktop/salvardor_articles/StateCapacity")

```
```{r}
library(WDI)
library(sqldf)
```
The WDI Package (Arel-Bundock 2013) provides an R wrapper for the World Development indicators API. The data we collect from the WDI site include the population total, the number of homicides per million residents, the population growth per year, the GDP Growth per year, and GNI per capita, per year. The sqldf package (Grothendieck 2014) is used to execute arbitrary SQL queries inside of the native R environment. 
```{r}
#### Collect the Relevant WDI Indicaters #### 
# "SP.POP.TOTL" - Total Population
# "VC.IHR.PSRC.P5" - Homicides "VC.IHR.PSRC.P5"
#  "SP.POP.GROW" - PopulationGrowth 
# "NY.GDP.MKTP.KD.ZG- GDPGrowth
# "NY.GNP.PCAP.CD GNIPC - GNI Percapita 

data <- WDI(country = "all", indicator = c("SP.POP.TOTL","VC.IHR.PSRC.P5",
                                           "SP.POP.GROW","NY.GDP.MKTP.KD.ZG","NY.GNP.PCAP.CD"),
            start = 1990, end = 2015, extra = FALSE, cache = NULL)
```
To prepare the data for the regression analysis, we first rename the columns return by the WDI function to more intuitive names.  Then, we select out only the three countries were are interested in. Next, we append the data for alien apprehensions to the resulting data frame. Finally, we normalize the 
apprehension data by population.  

```{r}
cnames <- c("Code","Country","Year","Population","Homicides","PopulationGrowth",
              "GDPGrowth","GNIPC")
colnames(data) <- cnames


TriangleData <- sqldf("SELECT * FROM data WHERE (Country = 'El Salvador'
                      OR Country = 'Guatemala' OR Country = 'Honduras')
                      AND (Year > 2001 AND Year < 2014)
                      ORDER BY Country,Year ASC")

aliens <- read.csv("aliens.csv")
TriangleData$Aliens <- aliens$Aliens
TriangleData$OneHundredK <-  (TriangleData$Population / 100000)
TriangleData$Aliens100K <- (TriangleData$Aliens / TriangleData$OneHundredK)
```

We use the head() function to provide a sample of the dataset. 
```{r}
head(TriangleData)
```
There's a NULL Value for 2013 in Guatemala. I am going to do a simple impute of the data by setting the value the same as 2012. 
```{r}
TriangleData[24,]$Homicides <- 34.57425

```
### Regression Model 
The lm() function in the base R package provides a set of tools to build and evaluate a linear model. In this model, we use apprehension data normalized by population as the response variable. The model coefficients are the population growth, GDP growth, GNI per capita, and the data for homicides per million residents, per year.   

To test for a relationship, we can use a t-test to examine whether or not the slope is affected by the homicide variable.  We set the null hypothesis that the slope is zero and the alternative hypothesis that the slope between the dependent variable and the homicides coefficient is greater than zero. We will reject the hypothesis if the results indicate a p-value less than .05. 

```{r}
emigrationModel <- lm(TriangleData$Aliens100K ~ TriangleData$PopulationGrowth
                       + TriangleData$GDPGrowth + TriangleData$GNIPC + TriangleData$Homicides)

summary(emigrationModel)

```
The model summary above supports the conclusion that a statistically significant relationship exists between increased violence in the Northern Triangle and eventual migration to the United States. 
The p-test of slope for the homicides variable is significant at a .05 significance level. This gives us reasonable support to reject the null hypothesis that there is no relationship between the homicide variable and the number of migrants. As a result, we can conclude that a relationship exists between increased homicides and increased migration to the United States. 

The Adjusted R-Squared value shows how well the model explains the variance in the data.  The model explains approximately 37% of the variance in the model. As discussed above, additional variance like a change in American immigration policy between years is not accounted for in this model. However, this analysis supports the general statement that migration to the United States increases as homicides in Honduras, El Salvador, and Guatemala increase. Furthermore, the F-statistic and corresponding p-value support the statement that our model outperforms random noise. 

It is interesting to note that emigration from the Northern Triangle increases over time even though the relative economic situation of all three countries improves as measured by GDP Growth and GNI per capita. This supports our argument that emigration from the Triangle is more nuanced than economic security and is partially determined by physical security. 


```{r}
testData <- sqldf("SELECT PopulationGrowth,GDPGrowth,GNIPC,Homicides
                  FROM TriangleData")
predictedAliens <- predict(emigrationModel,testData$OneHundredK)
```
The plot below shows the performance of the model as determined by the prediction plot. On the Y-axis, we plot the values our model predicted for our dataset. On the X-axis, we plot the actual values for our data set. If the model predicts the data set perfectly, there will be a diagonal line drawn along the X and Y axis. 


```{r,eco=FALSE}
plot(predictedAliens,TriangleData$Aliens100K,
     main="Predicted vs Detentions per Capita",
     xlab="Actual",
     ylab="Predicted")
```


The plot above illustrates that model does not perfectly forecast the number of aliens apprehended. However, a clear trend is visible and illustrates the utility of the model for anticipating increased migration based on deteriorating physical security within El Salvador, Guatemala, and Honduras. 


The plot below shows the trend in internal security and emigration to the United States over time for all three countries in the Northern Triangle. The X-axis shows the number of aliens apprehended per 100,000 residents and the Y-axis shows the number of homicides that occurred per million residents. The animation shows the change in values over time. The radius of the three circles reflects the GNI per capita. As illustrated by the visualization, the GNI increases for each nation over time. You can read explicit values for each country by hovering over the circle. Finally, it is possible to control the view of each year by hovering over the year counter and dragging it to the desired year. 

You can open the visualization in it's own page by clicking <a href="https://morning-sea-82112.herokuapp.com/immigration"> here. </a>

<iframe style="border-style: none;width: 100%; height: 500px;" src="https://morning-sea-82112.herokuapp.com/immigration">  </iframe>

### Works Cited 
Arel-Bundock, Vincent. "WDI: World Development Indicators (World Bank)." R package version 2 (2013).

Grothendieck, Gabor. "sqldf: perform SQL selects on R data frames." (2012).

Office of Immigration Statistics, US Dep't of Homeland Security,. "2013 Yearbook of Immigration Statistics." (2014): 20.

World Bank Group, ed. World Development Indicators 2016. World Bank Publications, 2016.